---
title: "Differential Expression & Gene Set Analysis – COPD (GSE57148)"
author: "Based on Kim et al., 2015 – Int J Genomics"
output:
  html_document:
    theme: cerulean
    toc: yes
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 9, fig.height = 6,
                      fig.path = "fig/DGE-", warning = FALSE, message = FALSE)
```

# Introduction

In this tutorial we identify differentially expressed genes (DEGs) between
**COPD** and **Control** lung tissue samples (GSE57148, Kim et al. 2015) and
then perform gene set analysis to identify the biological pathways that are
dysregulated in COPD.

### Paper strategy

The paper used two approaches:

1. **Differential expression**: Student's *t*-test on FPKM-normalized values
   (upper-quartile normalized), with FDR correction using the
   Benjamini–Hochberg algorithm (*q* < 0.01). They identified **2,312 DEGs**.

2. **Gene set analysis**:
   - **DAVID** (v6.7) — ORA-style functional annotation of DEGs (GO terms)
   - **GSEA** (v2.0.8) with **MSigDB** (v3.1) — ranked gene-set enrichment
     using all quantified genes ranked by log fold-change
   - Key pathways found: **oxidative phosphorylation** (down), **protein
     catabolism** (down), **chromatin modification** (up)

We reproduce both approaches using modern Bioconductor tools:

- **edgeR GLM** (negative binomial model on raw counts) — more statistically
  principled than the paper's t-test on normalized values, but we also run the
  **t-test on FPKM-UQ** to directly mirror the paper
- **DESeq2** *(new package)* — a second NB-based DE method with a different
  dispersion shrinkage strategy; used as cross-validation against edgeR
- **NOISeq** *(new package)* — a non-parametric DE method that makes no
  distributional assumptions; uniquely suited to detect DE without relying on
  the NB model
- **clusterProfiler** for ORA (KEGG, Reactome, GO) — equivalent to DAVID
- **GSEA** via `clusterProfiler::gseKEGG` / `gseGO` — equivalent to the
  paper's GSEA with MSigDB

**Files required:**

- `normalization.RData` (output of previous tutorial)

---

## Package installation

```{r, eval=FALSE}
install.packages(c("ggplot2", "ggrepel", "pheatmap", "cluster"))
if (!require("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install(c("edgeR", "limma", "EDASeq", "genefilter",
                       "clusterProfiler", "org.Hs.eg.db", "ReactomePA",
                       "enrichplot", "pathview", "DESeq2", "NOISeq",
                       "AnnotationDbi", "ggnewscale", "ggupset", "ggridges"))
```

## Load packages

```{r, message=FALSE}
library(edgeR)
library(limma)
library(EDASeq)
library(DESeq2)
library(NOISeq)
library(genefilter)
library(ggplot2)
library(ggrepel)
library(pheatmap)
library(cluster)
library(clusterProfiler)
library(org.Hs.eg.db)
library(ReactomePA)
library(enrichplot)
library(apeglm)
library(AnnotationDbi)
library(ggridges)
library(dplyr)
```

## Load data from previous tutorial

```{r}
load("normalization.RData")

# Confirm objects available
cat("Objects loaded:", paste(ls(), collapse = ", "), "\n")
cat("raw_count dimensions:", dim(raw_count), "\n")
cat("Group breakdown:\n")
print(table(group))
```

---

# Mean–Variance Relationship

Before running the DE test we examine the mean–variance relationship. In
RNA-seq data the variance is larger than the mean (**overdispersion**), which
is why the **Negative Binomial (NB)** distribution is used instead of a
Poisson model.

$$\text{Var}(Y) = \mu + \phi\mu^2$$

where $\phi$ is the dispersion parameter capturing extra-Poisson variability.

```{r}
# Design matrix: COPD vs Control
design <- model.matrix(~ group)
head(design)

# Build DGEList and compute TMM normalization factors
dge <- DGEList(counts = raw_count, group = group)
dge <- calcNormFactors(dge, method = "TMM")

cat("TMM normalization factors (first 6):\n")
print(round(dge$samples$norm.factors[1:6], 4))
```

```{r}
# Estimate dispersions (common → trended → tagwise)
dge <- estimateDisp(dge, design, robust = TRUE)

cat("Common dispersion:", round(dge$common.dispersion, 4), "\n")
cat("Biological coefficient of variation (BCV):",
    round(sqrt(dge$common.dispersion), 4), "\n")
```

```{r}
# Mean–variance plot
plotMeanVar(dge,
            show.raw.vars     = TRUE,
            show.ave.raw.vars = FALSE,
            NBline            = TRUE,
            show.tagwise.vars = TRUE,
            main = "Mean–Variance relationship (COPD dataset)")
```

> Most points lie above the Poisson line (black), confirming overdispersion.
> The NB line (blue) fits the data much better, justifying the NB model.

```{r}
# BCV plot — shows per-gene dispersion estimates
plotBCV(dge, main = "Biological Coefficient of Variation")
```

---

# Differential Expression Analysis

## Approach 1 — edgeR GLM (NB model on raw counts)

This is the modern recommended approach. edgeR fits a negative-binomial
generalised linear model and uses a likelihood ratio test (LRT) to identify
DEGs.

```{r}
# Fit the GLM
fit <- glmFit(dge, design)

# Likelihood ratio test: COPD vs Control (coef = 2)
res <- glmLRT(fit, coef = 2)

# All results table
tab_edger <- topTags(res, n = Inf)$table
tab_edger$gene <- rownames(tab_edger)

cat("Total genes tested:", nrow(tab_edger), "\n")
```

```{r}
# Apply paper's FDR threshold: q < 0.01
de_edger <- tab_edger[tab_edger$FDR < 0.01, ]
cat("DEGs at FDR < 0.01 (edgeR):", nrow(de_edger), "\n")
cat("  Up-regulated in COPD:", sum(de_edger$logFC > 0), "\n")
cat("  Down-regulated in COPD:", sum(de_edger$logFC < 0), "\n")
```

```{r}
cat("\nTop 20 DEGs by FDR:\n")
head(tab_edger[order(tab_edger$FDR), ], 20)
```

## Approach 2 — Student's t-test on FPKM-UQ (paper method)

The paper applied a Student's t-test directly on the FPKM upper-quartile
normalized values, then corrected for multiple testing with
Benjamini–Hochberg. We reproduce this to check how closely we can match the
paper's 2,312 DEGs.

```{r}
# fpkm_uq was saved in normalization.RData
# Apply log1p transformation to stabilize variance before t-test
log_fpkm_uq <- log1p(fpkm_uq)

ctrl_idx <- which(group == "Control")
copd_idx <- which(group == "COPD")

# Run t-test for each gene
pvals <- apply(log_fpkm_uq, 1, function(x) {
  t.test(x[copd_idx], x[ctrl_idx], var.equal = FALSE)$p.value
})

# Compute log fold-change (mean COPD - mean Control on log scale)
lfc_ttest <- rowMeans(log_fpkm_uq[, copd_idx]) -
             rowMeans(log_fpkm_uq[, ctrl_idx])

# Benjamini-Hochberg FDR correction
qvals <- p.adjust(pvals, method = "BH")

tab_ttest <- data.frame(
  gene   = rownames(log_fpkm_uq),
  logFC  = lfc_ttest,
  PValue = pvals,
  FDR    = qvals,
  stringsAsFactors = FALSE
)
tab_ttest <- tab_ttest[order(tab_ttest$FDR), ]

de_ttest <- tab_ttest[tab_ttest$FDR < 0.01, ]
cat("DEGs at FDR < 0.01 (t-test on FPKM-UQ):", nrow(de_ttest), "\n")
cat("  Paper reported: 2,312 DEGs\n")
cat("  Up-regulated in COPD:", sum(de_ttest$logFC > 0), "\n")
cat("  Down-regulated in COPD:", sum(de_ttest$logFC < 0), "\n")
```

```{r}
# Overlap between the two methods
overlap <- intersect(de_edger$gene, de_ttest$gene)
cat("\nDEGs found by both methods:", length(overlap), "\n")
cat("DEGs found only by edgeR:", sum(!de_edger$gene %in% de_ttest$gene), "\n")
cat("DEGs found only by t-test:", sum(!de_ttest$gene %in% de_edger$gene), "\n")
```

> For downstream gene set analyses we use both lists but focus primarily on
> the **t-test result** to directly reproduce the paper, and use the **edgeR
> result** as a modern cross-validation.

---

# DEG Exploration

## MD plot

The MD (Mean–Difference) plot shows log fold-change vs average log expression.
Red = up-regulated in COPD, blue = down-regulated, black = not significant.

```{r}
plotMD(res,
       main = "MD plot – COPD vs Control (edgeR)",
       status = ifelse(tab_edger$FDR < 0.01,
                       ifelse(tab_edger$logFC > 0, "Up", "Down"),
                       "NS"),
       values = c("Up", "Down", "NS"),
       col    = c("red", "blue", "grey70"),
       legend = "topright")
abline(h = 0, lty = 2, col = "grey40")
```

## Volcano plot

```{r, fig.width=10, fig.height=8}
# Safe symbol lookup using the annotation data frame directly
tab_edger$symbol <- annot_sub$Symbol[match(tab_edger$gene, rownames(raw_count))]

# Replace NA or empty symbols with the Entrez ID so labels are never blank
tab_edger$symbol[is.na(tab_edger$symbol) | tab_edger$symbol == ""] <- 
  tab_edger$gene[is.na(tab_edger$symbol) | tab_edger$symbol == ""]

tab_edger$direction <- "NS"
tab_edger$direction[tab_edger$FDR < 0.01 & tab_edger$logFC > 0] <- "Up (COPD)"
tab_edger$direction[tab_edger$FDR < 0.01 & tab_edger$logFC < 0] <- "Down (COPD)"

# Top genes to label (by FDR)
top_label <- head(tab_edger[order(tab_edger$FDR), ], 30)

ggplot(tab_edger, aes(x = logFC, y = -log10(PValue), colour = direction)) +
  geom_point(size = 0.8, alpha = 0.6) +
  scale_colour_manual(values = c("Up (COPD)"   = "#E05C5C",
                                 "Down (COPD)" = "#5B8DB8",
                                 "NS"          = "grey70")) +
  geom_text_repel(data = top_label,
                  aes(label = symbol),
                  size = 2.5, max.overlaps = 25,
                  colour = "black") +
  geom_hline(yintercept = -log10(max(tab_edger$PValue[tab_edger$FDR < 0.01])),
             lty = 2, colour = "grey40") +
  geom_vline(xintercept = c(-1, 1), lty = 2, colour = "grey40") +
  labs(title  = "Volcano plot – COPD vs Control",
       x      = expression(log[2]~Fold~Change),
       y      = expression(-log[10]~p-value),
       colour = "Direction") +
  theme_classic(base_size = 13)
```

## P-value distribution

In an unbiased experiment we expect a spike at low p-values mixed with a
uniform background. A predominantly flat or inverse distribution suggests a
problematic test; a clean spike indicates genuine signal.

```{r}
hist(tab_edger$PValue,
     breaks = 50, col = "#5B8DB8", border = "white",
     main = "P-value distribution – edgeR (COPD vs Control)",
     xlab = "p-value")
```

```{r}
hist(tab_ttest$PValue,
     breaks = 50, col = "#E05C5C", border = "white",
     main = "P-value distribution – t-test on FPKM-UQ (paper method)",
     xlab = "p-value")
```

---

# Cluster Analysis

## Heatmap

We select only the DEGs (using the t-test list to match the paper) and
visualize them on the TMM-normalized matrix. The `scale="row"` option centres
and scales each gene across samples so that relative expression patterns are
easy to compare.

```{r, fig.width=12, fig.height=9}
# Use TMM counts for visualization (robust normalized)
deg_genes_ttest <- de_ttest$gene

# Subset to DEG genes
heatmap_mat <- log1p(tmm[rownames(tmm) %in% deg_genes_ttest, ])

# Add gene symbols as row names where available
sym_map <- annot_sub$symbol[match(rownames(heatmap_mat), rownames(raw_count))]
rownames(heatmap_mat) <- ifelse(is.na(sym_map) | sym_map == "",
                                rownames(heatmap_mat), sym_map)

# Column annotation
col_ann <- data.frame(Group = group)
rownames(col_ann) <- colnames(heatmap_mat)

ann_colors <- list(Group = c("Control" = "#5B8DB8", "COPD" = "#E05C5C"))

g <- pheatmap(heatmap_mat,
              show_colnames  = FALSE,
              show_rownames  = FALSE,
              scale          = "row",
              annotation_col = col_ann,
              annotation_colors = ann_colors,
              clustering_method = "complete",
              main = "Heatmap – DEGs (FDR < 0.01, t-test on FPKM-UQ)",
              silent = TRUE)
print(g)
```

## Silhouette index

We cut the sample dendrogram into 2 clusters (COPD and Control) and evaluate
how well the clustering captures the known biological groups using the
**silhouette index**.

$$s(i) = \frac{b(i) - a(i)}{\max\{a(i),\, b(i)\}}$$

Values near +1 = well-clustered; near 0 = uncertain; negative = likely
misclassified.

```{r}
cut      <- cutree(g$tree_col, k = 2)
dist_mat <- dist(t(heatmap_mat))
sil      <- silhouette(cut, dist_mat)
plot(sil,
     col    = c("#5B8DB8", "#E05C5C"),
     border = NA,
     main   = "Silhouette plot – DEG-based clustering")
cat("Average silhouette width:", round(mean(sil[, 3]), 3), "\n")
```

---

# Gene Set Analysis

Gene set analysis answers the question: **which biological processes and
pathways are enriched among the DEGs?** We use two complementary approaches:

| Approach | Method | Paper equivalent |
|---|---|---|
| **ORA** | Fisher exact test on DEG list | DAVID v6.7 |
| **GSEA** | Ranked enrichment on all genes | GSEA v2.0.8 + MSigDB |

## Prepare gene identifiers

Our row names are **Entrez Gene IDs** — the native format for KEGG, Reactome,
and GO databases. No ID conversion is needed.

```{r}
# Universe = all tested genes (Entrez IDs as character)
universe_entrez <- rownames(tab_ttest)

# Significant DEGs
sig_entrez <- de_ttest$gene

# Ranked list for GSEA: all genes sorted by logFC (t-test), descending
ranked_list        <- tab_ttest$logFC
names(ranked_list) <- tab_ttest$gene
ranked_list        <- sort(ranked_list[!duplicated(names(ranked_list))],
                           decreasing = TRUE)

cat("Universe size:", length(universe_entrez), "\n")
cat("DEG list size:", length(sig_entrez), "\n")
cat("Ranked list size:", length(ranked_list), "\n")
```

---

## ORA — Over-Representation Analysis

ORA tests whether a gene set is statistically over-represented in the DEG
list compared to the universe (all tested genes), using a Fisher exact test.

### GO Biological Process

The paper used DAVID to identify GO terms enriched in the DEGs. The top
categories reported were:

- **Oxidative phosphorylation** (down)
- **Protein catabolism / proteasome** (down)
- **Chromatin modification** (up)
- **Ribonucleoprotein complex** (up)

```{r}
ego_BP <- enrichGO(
  gene          = sig_entrez,
  universe      = universe_entrez,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENTREZID",
  ont           = "BP",
  pvalueCutoff  = 0.05,
  pAdjustMethod = "BH",
  readable      = TRUE   # converts Entrez IDs to gene symbols in output
)

cat("Enriched GO-BP terms:", nrow(ego_BP@result[ego_BP@result$p.adjust < 0.05, ]),
    "\n\n")
head(ego_BP@result[, 1:7], 15)
```

```{r, fig.width=12, fig.height=7}
barplot(ego_BP, showCategory = 20,
        title = "GO Biological Process – ORA (COPD DEGs)")
```

```{r, fig.width=12, fig.height=7}
dotplot(ego_BP, showCategory = 20,
        title = "GO-BP dotplot")
```

```{r, fig.width=14, fig.height=8}
cnetplot(ego_BP, showCategory = 10) +
  ggtitle("GO-BP network – gene–term connections")
```

```{r, fig.width=10, fig.height=7}
ego_BP_sim <- pairwise_termsim(ego_BP)
emapplot(ego_BP_sim, showCategory = 30) +
  ggtitle("GO-BP enrichment map – term similarity")
```

### GO Molecular Function

```{r}
ego_MF <- enrichGO(
  gene          = sig_entrez,
  universe      = universe_entrez,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENTREZID",
  ont           = "MF",
  pvalueCutoff  = 0.05,
  pAdjustMethod = "BH",
  readable      = TRUE
)
cat("Enriched GO-MF terms:", nrow(ego_MF@result[ego_MF@result$p.adjust < 0.05, ]), "\n")
```

```{r, fig.width=12, fig.height=7}
if (nrow(ego_MF@result[ego_MF@result$p.adjust < 0.05, ]) > 0) {
  barplot(ego_MF, showCategory = 20, title = "GO Molecular Function – ORA")
}
```

### GO Cellular Component

```{r}
ego_CC <- enrichGO(
  gene          = sig_entrez,
  universe      = universe_entrez,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENTREZID",
  ont           = "CC",
  pvalueCutoff  = 0.05,
  pAdjustMethod = "BH",
  readable      = TRUE
)
cat("Enriched GO-CC terms:", nrow(ego_CC@result[ego_CC@result$p.adjust < 0.05, ]), "\n")
```

```{r, fig.width=12, fig.height=7}
if (nrow(ego_CC@result[ego_CC@result$p.adjust < 0.05, ]) > 0) {
  barplot(ego_CC, showCategory = 20, title = "GO Cellular Component – ORA")
}
```

### KEGG pathways

```{r}
kk_ora <- enrichKEGG(
  gene          = sig_entrez,
  universe      = universe_entrez,
  organism      = "hsa",
  keyType       = "ncbi-geneid",
  pvalueCutoff  = 0.05,
  pAdjustMethod = "BH"
)

cat("Enriched KEGG pathways:", nrow(kk_ora@result[kk_ora@result$p.adjust < 0.05, ]),
    "\n\n")
head(kk_ora@result[, 1:7], 15)
```

```{r, fig.width=12, fig.height=7}
barplot(kk_ora, showCategory = 20,
        title = "KEGG pathways – ORA (COPD DEGs)")
```

```{r, fig.width=12, fig.height=7}
dotplot(kk_ora, showCategory = 20, title = "KEGG dotplot")
```

```{r, fig.width=14, fig.height=8}
kk_ora_sim <- pairwise_termsim(kk_ora)
emapplot(kk_ora_sim, showCategory = 30) +
  ggtitle("KEGG enrichment map")
```

```{r, fig.width=12, fig.height=8}
heatplot(kk_ora, showCategory = 15) +
  ggtitle("KEGG heatplot – gene–pathway membership")
```

### Reactome pathways

```{r}
rp_ora <- enrichPathway(
  gene          = sig_entrez,
  universe      = universe_entrez,
  organism      = "human",
  pvalueCutoff  = 0.05,
  pAdjustMethod = "BH",
  readable      = TRUE
)

cat("Enriched Reactome pathways:",
    nrow(rp_ora@result[rp_ora@result$p.adjust < 0.05, ]), "\n\n")
head(rp_ora@result[, 1:7], 15)
```

```{r, fig.width=12, fig.height=7}
if (nrow(rp_ora@result[rp_ora@result$p.adjust < 0.05, ]) > 0) {
  barplot(rp_ora, showCategory = 20, title = "Reactome pathways – ORA")
  dotplot(rp_ora, showCategory = 20, title = "Reactome dotplot")
}
```

---

## GSEA — Gene Set Enrichment Analysis

GSEA does not require a binary DEG threshold. All quantified genes are ranked
by log fold-change (COPD vs Control), and the algorithm tests whether members
of a gene set cluster at the top (up in COPD) or bottom (down in COPD) of
this ranking beyond what is expected by chance.

This directly reproduces the paper's GSEA analysis (they used MSigDB v3.1;
here we use the current KEGG and GO databases through clusterProfiler).

The key output is the **Normalized Enrichment Score (NES)**:

- **NES > 0** → gene set enriched at the top → genes are **up-regulated in COPD**
- **NES < 0** → gene set enriched at the bottom → genes are **down-regulated in COPD**

### GSEA — KEGG pathways

```{r}
set.seed(42)   # for reproducibility of permutation p-values
kk_gsea <- gseKEGG(
  geneList     = ranked_list,
  organism     = "hsa",
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH",
  verbose      = FALSE
)

cat("GSEA KEGG — enriched pathways:", nrow(kk_gsea@result), "\n\n")
# Show top up- and down-regulated pathways
res_kegg <- kk_gsea@result[order(kk_gsea@result$NES, decreasing = TRUE), ]
cat("Top 10 pathways enriched in COPD (NES > 0):\n")
print(head(res_kegg[res_kegg$NES > 0, c("Description","NES","p.adjust")], 10))
cat("\nTop 10 pathways depleted in COPD (NES < 0):\n")
print(head(res_kegg[res_kegg$NES < 0, c("Description","NES","p.adjust")], 10))
```

```{r, fig.width=14, fig.height=8}
dotplot(kk_gsea, showCategory = 20, split = ".sign") +
  facet_grid(. ~ .sign) +
  ggtitle("KEGG GSEA – COPD vs Control")
```

```{r, fig.width=12, fig.height=7}
ridgeplot(kk_gsea, showCategory = 20, label_format = 50) +
  ggtitle("KEGG GSEA – logFC distribution per gene set")
```

```{r}
# GSEA enrichment plot for the top down-regulated pathway
# Oxidative phosphorylation is expected to be the most depleted (paper finding)
oxphos_id <- res_kegg$ID[grepl("oxidative|phosphorylation|OXPHOS",
                                res_kegg$Description,
                                ignore.case = TRUE)][1]
if (!is.na(oxphos_id)) {
  gseaplot(kk_gsea, geneSetID = oxphos_id,
           title = paste0("GSEA plot – ", oxphos_id,
                          " (", res_kegg$Description[res_kegg$ID == oxphos_id][1], ")"))
}
```

```{r}
# GSEA enrichment plot for the top up-regulated pathway
top_up_id <- res_kegg$ID[res_kegg$NES > 0][1]
if (!is.na(top_up_id)) {
  gseaplot(kk_gsea, geneSetID = top_up_id,
           title = paste0("GSEA plot – ", top_up_id,
                          " (", res_kegg$Description[res_kegg$ID == top_up_id][1], ")"))
}
```

### GSEA — GO Biological Process

```{r}
set.seed(42)
kkGO_gsea <- gseGO(
  geneList      = ranked_list,
  ont           = "BP",
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENTREZID",
  pvalueCutoff  = 0.05,
  pAdjustMethod = "BH",
  verbose       = FALSE
)

cat("GSEA GO-BP — enriched terms:", nrow(kkGO_gsea@result), "\n\n")
res_go <- kkGO_gsea@result[order(kkGO_gsea@result$NES, decreasing = TRUE), ]
cat("Top 10 GO-BP terms enriched in COPD (NES > 0):\n")
print(head(res_go[res_go$NES > 0, c("Description","NES","p.adjust")], 10))
cat("\nTop 10 GO-BP terms depleted in COPD (NES < 0):\n")
print(head(res_go[res_go$NES < 0, c("Description","NES","p.adjust")], 10))
```

```{r, fig.width=14, fig.height=8}
dotplot(kkGO_gsea, showCategory = 20, split = ".sign") +
  facet_grid(. ~ .sign) +
  ggtitle("GO-BP GSEA – COPD vs Control")
```

```{r, fig.width=12, fig.height=7}
ridgeplot(kkGO_gsea, showCategory = 20, label_format = 50) +
  ggtitle("GO-BP GSEA – logFC distribution per gene set")
```

```{r}
# Enrichment plot for oxidative phosphorylation (paper's top finding)
oxphos_go <- res_go$ID[grepl("oxidative phosphorylation",
                              res_go$Description,
                              ignore.case = TRUE)][1]
if (!is.na(oxphos_go)) {
  gseaplot(kkGO_gsea, geneSetID = oxphos_go,
           title = paste0("GSEA plot – ",
                          res_go$Description[res_go$ID == oxphos_go][1]))
}
```

```{r}
# Enrichment plot for proteasome / protein catabolic process (paper's second finding)
prot_go <- res_go$ID[grepl("proteaso|protein catabol|ubiquitin",
                             res_go$Description,
                             ignore.case = TRUE)][1]
if (!is.na(prot_go)) {
  gseaplot(kkGO_gsea, geneSetID = prot_go,
           title = paste0("GSEA plot – ",
                          res_go$Description[res_go$ID == prot_go][1]))
}
```

---

# New Package 1 — DESeq2

## Why DESeq2?

**DESeq2** (Love et al., 2014) is an alternative to edgeR for DE analysis of
RNA-seq count data. Both packages use the negative binomial distribution, but
they differ in how they estimate and stabilize the per-gene dispersion:

| Feature | edgeR | DESeq2 |
|---|---|---|
| Dispersion model | Empirical Bayes, tagwise | Shrinkage via MAP estimation |
| Normalization | TMM scaling factors | Median-of-ratios (size factors) |
| Test | Likelihood Ratio Test (LRT) | Wald test (or LRT) |
| Fold-change shrinkage | Optional (apeglm) | Built-in (lfcShrink) |

Running both gives a robust cross-validation: genes identified as DE by both
methods are far more likely to be genuine findings than those found by only one.

```{r}
# DESeq2 requires integer counts and a design DataFrame
dds <- DESeqDataSetFromMatrix(
  countData = round(raw_count),          # must be integers
  colData   = DataFrame(group = group),
  design    = ~ group
)

# Remove very low-count genes (pre-filter recommended by DESeq2 vignette)
dds <- dds[rowSums(counts(dds)) >= 10, ]
cat("Genes after pre-filter:", nrow(dds), "\n")
```

```{r}
# Run the full DESeq2 pipeline:
#   1. Estimate size factors (median-of-ratios normalization)
#   2. Estimate dispersions with shrinkage
#   3. Fit negative binomial GLM and Wald test
dds <- DESeq(dds)

cat("Size factors (first 6 samples):\n")
print(round(sizeFactors(dds)[1:6], 4))
```

### DESeq2 dispersion plot

DESeq2 estimates a per-gene dispersion (grey dots), fits a trend (red line),
and then shrinks each gene's dispersion towards the trend (blue dots). Genes
with very high dispersion relative to the trend are not shrunk (circled black
dots — these are treated as outliers to avoid inflating false positives).

```{r}
plotDispEsts(dds,
             main = "DESeq2 – Dispersion estimates (COPD vs Control)")
```

### DESeq2 results

```{r}
# Extract results: COPD vs Control
# contrast = c("factor", "numerator", "denominator")
res_deseq2 <- results(dds,
                      contrast       = c("group", "COPD", "Control"),
                      alpha          = 0.01,
                      pAdjustMethod  = "BH")

summary(res_deseq2)
```

```{r}
# Apply log fold-change shrinkage (apeglm) for better effect size estimates
# This reduces noise for low-count genes without changing significance calls
res_deseq2_shrunk <- lfcShrink(dds,
                                coef    = "group_COPD_vs_Control",
                                type    = "apeglm",
                                res     = res_deseq2)
```

```{r}
# Convert to data frame and filter to DEGs
tab_deseq2 <- as.data.frame(res_deseq2_shrunk)
tab_deseq2$gene <- rownames(tab_deseq2)
tab_deseq2 <- tab_deseq2[!is.na(tab_deseq2$padj), ]

de_deseq2 <- tab_deseq2[tab_deseq2$padj < 0.01, ]
cat("DEGs at FDR < 0.01 (DESeq2):", nrow(de_deseq2), "\n")
cat("  Up-regulated in COPD:", sum(de_deseq2$log2FoldChange > 0), "\n")
cat("  Down-regulated in COPD:", sum(de_deseq2$log2FoldChange < 0), "\n")
```

### DESeq2 MA plot

The MA plot (built-in to DESeq2) shows the shrunken log fold-changes vs mean
expression. Grey points = not significant; blue points = significant DEGs.
The fold-change shrinkage pulls low-count genes' extreme estimates towards
zero, making the plot more informative than an unshrunken version.

```{r}
plotMA(res_deseq2_shrunk,
       alpha = 0.01,
       main  = "DESeq2 MA plot – COPD vs Control (shrunken LFC)",
       ylim  = c(-4, 4))
```

### DESeq2 volcano plot

```{r, fig.width=10, fig.height=8}
tab_deseq2$direction <- "NS"
tab_deseq2$direction[tab_deseq2$padj < 0.01 &
                     tab_deseq2$log2FoldChange > 0] <- "Up (COPD)"
tab_deseq2$direction[tab_deseq2$padj < 0.01 &
                     tab_deseq2$log2FoldChange < 0] <- "Down (COPD)"

tab_deseq2$symbol <- annot_sub$symbol[match(tab_deseq2$gene,
                                             rownames(raw_count))]
top_label_d2 <- head(tab_deseq2[order(tab_deseq2$padj, na.last = TRUE), ], 25)

ggplot(tab_deseq2, aes(log2FoldChange, -log10(pvalue), colour = direction)) +
  geom_point(size = 0.8, alpha = 0.6) +
  scale_colour_manual(values = c("Up (COPD)"   = "#E05C5C",
                                 "Down (COPD)" = "#5B8DB8",
                                 "NS"          = "grey70")) +
  geom_text_repel(data = top_label_d2,
                  aes(label = symbol),
                  size = 2.5, max.overlaps = 25, colour = "black") +
  geom_vline(xintercept = c(-1, 1), lty = 2, colour = "grey40") +
  labs(title  = "Volcano plot – DESeq2 (COPD vs Control)",
       x      = expression(log[2]~Fold~Change~(shrunken)),
       y      = expression(-log[10]~p-value),
       colour = "Direction") +
  theme_classic(base_size = 13)
```

### Three-way DEG overlap

```{r}
# Genes called DE by all three methods
overlap_all3 <- Reduce(intersect, list(de_edger$gene,
                                       de_ttest$gene,
                                       de_deseq2$gene))
cat("DEGs found by ALL THREE methods:", length(overlap_all3), "\n")
cat("  edgeR only:", sum(!de_edger$gene  %in% union(de_ttest$gene, de_deseq2$gene)), "\n")
cat("  DESeq2 only:", sum(!de_deseq2$gene %in% union(de_ttest$gene, de_edger$gene)), "\n")
cat("  t-test only:", sum(!de_ttest$gene  %in% union(de_edger$gene, de_deseq2$gene)), "\n")
cat("  edgeR ∩ DESeq2:", length(intersect(de_edger$gene, de_deseq2$gene)), "\n")
```

> Genes in the three-way intersection are the most robust DEG candidates —
> they survive all three different statistical frameworks.

---

# New Package 2 — NOISeq

## Why NOISeq?

**NOISeq** (Tarazona et al., 2011; updated 2015) takes a fundamentally
different approach to DE analysis compared to edgeR and DESeq2:

- **No distributional assumption** — it is non-parametric and does not assume
  that counts follow a negative binomial (or any other) distribution
- **Noise model** — it estimates the noise distribution empirically by
  comparing all pairs of samples *within* the same condition, then assesses
  whether the observed fold-change between conditions exceeds this noise level
- **Probability-based output** — instead of a p-value, each gene gets a
  probability $q$ that its expression change is genuine signal rather than
  noise; $q \geq 0.8$ is the recommended threshold (equivalent to a 20% FDR)

This makes NOISeq particularly valuable when the NB assumptions of edgeR/DESeq2
may be questionable, and as an independent cross-validation strategy.

### NOISeq method options

| Mode | When to use |
|---|---|
| `noiseq` | Biological replicates available (our case — 91 controls, 98 COPD) |
| `noiseqbio` | Optimized for biological replicates with technical variance estimation |

We use **`noiseqbio`** because we have biological replicates and want the most
robust estimates.

```{r}
# NOISeq requires a NOISeq-format data object
# Build factor data frame for sample groups
sample_factors <- data.frame(
  group = factor(group, levels = c("Control", "COPD")),
  row.names = colnames(raw_count)
)

# Create NOISeq data object
# We supply the raw counts and the sample annotation
# Gene length (for within-sample normalization) is taken from annot_sub
gene_lengths_noiSeq <- rowData(se)$length
names(gene_lengths_noiSeq) <- rownames(raw_count)

noiseq_data <- readData(
  data        = raw_count,
  factors     = sample_factors,
  length      = gene_lengths_noiSeq   # used for RPKM within-sample normalization
)

cat("NOISeq data object created.\n")
print(noiseq_data)
```

```{r}
# Run noiseqbio
# norm = "rpkm": within-sample normalization (corrects gene length + depth)
#         equivalent to FPKM for single-end; matches paper's within-sample step
# k = 0.5: pseudocount to handle zeros
# nclust = 15: number of clusters for noise estimation
# r = 20: number of simulations
set.seed(42)
noiseq_res <- noiseqbio(
  input   = noiseq_data,
  norm    = "rpkm",
  factor  = "group",
  conditions = list(Control = which(group == "Control"),
                    COPD    = which(group == "COPD")),
  k       = 0.5,
  nclust  = 15,
  r       = 20,
  plot    = FALSE
)

cat("NOISeq analysis complete.\n")
```

```{r}
# Extract results
tab_noiSeq <- noiseq_res@results[[1]]
tab_noiSeq$gene <- rownames(tab_noiSeq)

# NOISeq uses probability q: q >= 0.8 corresponds to ~80% posterior probability
# of being genuinely differentially expressed (equivalent to FDR ~ 0.2)
# For a stricter threshold matching the paper, use q >= 0.9 (~FDR ~ 0.1)
de_noiSeq_08 <- tab_noiSeq[!is.na(tab_noiSeq$prob) & tab_noiSeq$prob >= 0.8, ]
de_noiSeq_09 <- tab_noiSeq[!is.na(tab_noiSeq$prob) & tab_noiSeq$prob >= 0.9, ]

cat("NOISeq DEGs at q >= 0.8 (~FDR 20%):", nrow(de_noiSeq_08), "\n")
cat("NOISeq DEGs at q >= 0.9 (~FDR 10%):", nrow(de_noiSeq_09), "\n")
cat("  Up-regulated in COPD (q >= 0.8):",
    sum(de_noiSeq_08$log2FC > 0, na.rm = TRUE), "\n")
cat("  Down-regulated in COPD (q >= 0.8):",
    sum(de_noiSeq_08$log2FC < 0, na.rm = TRUE), "\n")
```

### NOISeq probability plot

The NOISeq probability plot shows each gene's differential expression
probability ($q$) on the y-axis against its mean expression on the x-axis.
Genes with high $q$ (above the dashed line at 0.8) are called DE; the colour
indicates fold-change direction.

```{r}
DE.plot(noiseq_res,
        q    = 0.8,
        graphic = "expr",
        main = "NOISeq – Differential expression probability plot")
```

```{r}
# MD-style plot from NOISeq
DE.plot(noiseq_res,
        q    = 0.8,
        graphic = "MD",
        main = "NOISeq – MD plot (M = log2FC, D = mean expression)")
```

### NOISeq volcano-style plot

```{r, fig.width=10, fig.height=7}
tab_noiSeq$direction <- "NS"
tab_noiSeq$direction[!is.na(tab_noiSeq$prob) &
                     tab_noiSeq$prob >= 0.8 &
                     tab_noiSeq$log2FC > 0] <- "Up (COPD)"
tab_noiSeq$direction[!is.na(tab_noiSeq$prob) &
                     tab_noiSeq$prob >= 0.8 &
                     tab_noiSeq$log2FC < 0] <- "Down (COPD)"

tab_noiSeq$symbol <- annot_sub$symbol[match(tab_noiSeq$gene,
                                             rownames(raw_count))]

ggplot(tab_noiSeq[!is.na(tab_noiSeq$prob), ],
       aes(log2FC, prob, colour = direction)) +
  geom_point(size = 0.8, alpha = 0.6) +
  scale_colour_manual(values = c("Up (COPD)"   = "#E05C5C",
                                 "Down (COPD)" = "#5B8DB8",
                                 "NS"          = "grey70")) +
  geom_hline(yintercept = 0.8, lty = 2, colour = "grey40") +
  geom_vline(xintercept = c(-1, 1), lty = 2, colour = "grey40") +
  labs(title  = "NOISeq – Probability vs log2FC (COPD vs Control)",
       x      = expression(log[2]~Fold~Change),
       y      = "NOISeq probability (q)",
       colour = "Direction") +
  theme_classic(base_size = 13)
```

### Four-method DEG overlap

```{r}
# Consensus DEGs — found by at least 3 out of 4 methods
all_methods <- list(
  ttest   = de_ttest$gene,
  edgeR   = de_edger$gene,
  DESeq2  = de_deseq2$gene,
  NOISeq  = de_noiSeq_08$gene
)

# Count how many methods call each gene as DE
all_genes <- unique(unlist(all_methods))
method_counts <- sapply(all_genes, function(g) {
  sum(sapply(all_methods, function(m) g %in% m))
})

cat("DEGs by number of methods agreeing:\n")
print(table(method_counts))

# Consensus: called DE by >= 3 methods
consensus_degs <- names(method_counts)[method_counts >= 3]
cat("\nConsensus DEGs (≥ 3/4 methods agree):", length(consensus_degs), "\n")
```

```{r, fig.width=8, fig.height=5}
# Barplot of agreement
barplot(table(method_counts),
        col   = c("grey80", "#A8C8E8", "#5B8DB8", "#1B4F72"),
        xlab  = "Number of methods agreeing",
        ylab  = "Number of genes",
        main  = "Method agreement across 4 DE approaches",
        names.arg = paste0(names(table(method_counts)), " method(s)"))
```

> Genes called DE by **all four methods** represent the most conservative and
> robust set of DEGs — they are detected regardless of the statistical
> framework or normalization strategy used.

---

# Comparison with Paper Findings

The paper reported the following key results. Use this table to check whether
your results reproduce them.

| Paper finding | Paper result | Your result |
|---|---|---|
| Total DEGs (q < 0.01) | **2,312** | Check `nrow(de_ttest)` |
| edgeR DEGs (FDR < 0.01) | — | Check `nrow(de_edger)` |
| DESeq2 DEGs (FDR < 0.01) | — | Check `nrow(de_deseq2)` |
| NOISeq DEGs (q ≥ 0.8) | — | Check `nrow(de_noiSeq_08)` |
| Consensus DEGs (≥ 3/4 methods) | — | Check `length(consensus_degs)` |
| Top down-regulated pathway | **Oxidative phosphorylation** | Check GSEA NES < 0 |
| Top down-regulated process | **Protein catabolism / proteasome** | Check GSEA NES < 0 |
| Up-regulated process | **Chromatin modification** | Check GSEA NES > 0 |
| Up-regulated complex | **Ribonucleoprotein complex** | Check GO-CC / GO-BP |

> **Note on DEG count differences**: The paper computed FPKM using Cufflinks
> (which uses effective transcript lengths and alignment-aware quantification)
> while our FPKM uses genomic gene lengths from the annotation. Small
> differences in the DEG count are expected. The pathway-level conclusions
> should be robust.

---

# Discussion Questions

**Q1.** The paper used a Student's t-test on normalized values, while we also
ran edgeR, DESeq2, and NOISeq on raw counts. What is the key statistical
reason why count-based models are more appropriate than t-tests for RNA-seq
data?

**Q2.** edgeR and DESeq2 both use the negative binomial distribution but
estimate dispersion differently. Looking at the BCV plot (edgeR) and the
dispersion plot (DESeq2), how does each method handle genes with unreliable
dispersion estimates?

**Q3.** NOISeq uses a probability $q$ rather than a p-value. What does
$q \geq 0.8$ mean in terms of the noise model, and why is this approach
advantageous when the NB distributional assumption may not hold?

**Q4.** Looking at the four-method overlap barplot, how many genes are called
DE by all four methods? What does a large number of "1 method only" genes
tell you about the sensitivity vs specificity trade-off between the methods?

**Q5.** Looking at the p-value histogram, does the distribution look as
expected for a dataset with genuine differential expression? What would it
look like if there were no true signal?

**Q6.** The GSEA results rank pathways by their NES. Why is the NES more
informative than a raw enrichment score when comparing pathways of different
sizes?

**Q7.** The paper found oxidative phosphorylation to be the most
down-regulated pathway in COPD. Does your GSEA reproduce this finding? Is it
consistent across the ORA results as well?

---

# Save data for next tutorial

```{r}
save(list = c("tab_edger",   "tab_ttest",   "tab_deseq2",  "tab_noiSeq",
              "de_edger",    "de_ttest",    "de_deseq2",   "de_noiSeq_08",
              "consensus_degs",
              "sig_entrez", "universe_entrez",
              "ranked_list",
              "kk_gsea", "kkGO_gsea",
              "kk_ora", "ego_BP", "rp_ora",
              "group", "sample_meta"),
     file = "DEG.RData")

cat("Saved DEG.RData\n")
cat("  de_ttest      – DEGs by paper method (t-test, FDR < 0.01)\n")
cat("  de_edger      – DEGs by edgeR GLM (FDR < 0.01)\n")
cat("  de_deseq2     – DEGs by DESeq2 (FDR < 0.01, shrunken LFC)\n")
cat("  de_noiSeq_08  – DEGs by NOISeq (q >= 0.8)\n")
cat("  consensus_degs – DEGs called by >= 3/4 methods\n")
cat("  kk_gsea       – GSEA KEGG results\n")
cat("  kkGO_gsea     – GSEA GO-BP results\n")
```

---

# Session Information

```{r}
sessionInfo()
```
